---
title: "MBQN vignette"
#date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{MBQN Package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This package contains a modified quantile normalization (QN) for preprocessing and analysis of omics or other matrix-like organized data with intensity values biased by global, columnwise distortions of intensity mean and scale. The modification balances the mean intensity of features (rows) which are rank invariant (RI) or nearly rank invariant (NRI) across samples (columns) before quantile normalization [1]. This helps to prevent an over-correction of the intensity profiles of RI and NRI features by classical QN and therefore supports the reduction of systematics in downstream analyses. Additional package functions help to detect, identify, and visualize potential RI or NRI features in the data and demonstrate the use of the modification.

## Installation
To install this package, you need R version >= 3.6.

For installation from Bioconductor run in R:

```{r bc-installation, eval = FALSE}
# if (!requireNamespace("BiocManager", quietly = TRUE))
#    install.packages("BiocManager")
BiocManager::install("MBQN")
```

## Dependencies
The core of the MBQN package uses normalizeQuantiles() from the package `limma` [2], available at https://bioconductor.org/packages/release/bioc/html/limma.html, for computation of the quantile normalization. Optionally, `normalize.quantiles()` from the package preprocessCore [3], available at https://bioconductor.org/packages/release/bioc/html/preprocessCore.html, can be used. 

The function `getPXDfile()` in MBQN uses data from the PRIDE repository. To run this function one needs the R package `rpx` [4] to download the data. 

To install these packages in R run: 

```{r dependencies1, eval = FALSE}
# if (!requireNamespace("BiocManager", quietly = TRUE))
#    install.packages("BiocManager")
# BiocManager::install(pkgs = c("preprocessCore","limma","rpx","SummarizedExperiment"))
```


## Usage
The package provides two basic functions: `mbqn()` applies quantile 
normalization or mean/median-balanced quantile normalization to a matrix. 
`mbqnNRI()` applies quantile normalization and mean/median-balanced quantile 
normalization only to selected nearly rank invariant and rank invariant 
features, specified by a threshold or manually. The input matrix may contain 
NAs. To run one of these functions you will need to provide an input matrix 
(see Examples). The argument `FUN` is used to select between classical quantile normalization (default), and mean or median balanced quantile normalization. The function `mbqnGetNRIfeatures()` and `mbqnPlotRI()` can be used to check a data matrix for rank or nearly rank 
invariant features. They provide a list of potential RI/NRI features together 
with their rank invariance frequency and a graphical output. 

## Examples
Example 1: Generate a distorted omics-like matrix of log2-transformed 
intensities with missing values and a single rank invariant feature:
```{r example1, eval = TRUE}
## basic example
library("MBQN")
set.seed(1234)
# data generation 
mtx <- mbqnSimuData("omics.dep")
# data distortion
mtx <- mbqnSimuDistortion(mtx)$x.mod
```
```{r figure1, fig.height = 5, fig.width = 6, fig.align = "left", fig.cap = "Fig. 1 Boxplot of the unnormalized, distorted intensity data matrix. The first feature is an RI feature (red line). It has maximum intensity for each sample!"}
plot.new()
mbqnBoxplot(mtx, irow = 1, main = "Unnormalized")
```
<br />
Apply check for rank invariant (RI) or nearly rank invariant (NRI) features to 
the data matrix and visualize result:
```{r, eval = TRUE}
res <- mbqnGetNRIfeatures(mtx, low_thr = 0.5)
```
```{r figure2, fig.height = 3, fig.width = 6, fig.align = "left", fig.cap = "Fig. 2 Correctly detected and identified RI feature with a data coverage of 100% across samples."}
plot.new()
mbqnPlotRI(res)
```
<br />
Apply quantile normalization with and without balancing the RI feature and 
compare the intensity features: 
```{r figure3, fig.height = 5, fig.width = 6, fig.align = 'left', fig.cap = "Fig. 3 Quantile normalized intensities with balanced and unbalanced normalized RI feature. Classical quantile normalization suppresses any intensity variation of the RI feature, while the MBQN preserves its variation while reducing systematic batch effects!"}
plot.new()
mtx.norm <- mbqnNRI(x = mtx, FUN = median, verbose = FALSE) # MBQN
mtx.qn <- mbqnNRI(x = mtx, FUN = NULL, verbose = FALSE) # QN
mbqnBoxplot(mtx.norm, irow = res$ip, vals = data.frame(QN = mtx.qn[res$ip,]), main = "Normalized")
```

<br /><br />
Example 2: Visualize  the effect of normalization on
rank mixing and rank invariant intensity features by comparing the intensity distribution of unnormalized, quantile and mean/median-balanced quantile normalized data on a matrix where rows represent features, e.g. of protein
abundances/intensities, and columns represent samples. 

```{r example2, eval = TRUE}
## basic example
library("MBQN")
set.seed(1234)
mtx <- mbqnSimuData("omics.dep") 
# Alternatively: mtx <- matrix(
#            c(5,2,3,NA,2,4,1,4,2,3,1,4,6,NA,1,3,NA,1,4,3,NA,1,2,3),ncol=4)

# perform qn, median balanced qn, and qn with median balanced nri feature
qn_mtx <- mbqn(mtx,FUN=NULL, verbose = FALSE)
mbqn_mtx <- mbqn(mtx,FUN = "median", verbose = FALSE)
qn_nri_mtx <- mbqnNRI(mtx,FUN = "median", low_thr = 0.5, verbose = FALSE)

# check saturation i.e. for rank invariance
res <- mbqnGetNRIfeatures(mtx, low_thr = 0.5)

ylim <- range(mtx, na.rm =TRUE)
```
```{r figure4, fig.height = 5, fig.width = 6, fig.align = "left", fig.cap = "Fig. 4 "}
# create a boxplot for qn-data with nri features median balanced
plot.new()
mbqnBoxplot(qn_nri_mtx, irow = res$ip,
            add.leg = TRUE, main ="QN data with MBQN NRI feature",
            cex.lab = 1.2,
            cex = 1.2,
            ylim = ylim)
```
```{r figure5, fig.height = 5, fig.width = 6, fig.align = "left", fig.cap = "Fig. 5 "}
# create a boxplot for x
plot.new()
mbqnBoxplot(mtx,
            add.leg = FALSE, main = "Data",
            cex.lab = 1.2,
            cex = 1.2,
            ylim = ylim)
```
```{r figure6, fig.height = 5, fig.width = 6, fig.align = "left", fig.cap = "Fig. 6 "}
# create a boxplot for qn-data
plot.new()
mbqnBoxplot(qn_mtx, irow = res$ip,
            add.leg = FALSE, main = "QN data",
            cex.lab = 1.2,
            cex = 1.2,
            ylim = ylim)
```
```{r figure7, fig.height = 5, fig.width = 6, fig.align = "left", fig.cap = "Fig. 7 "}
# create a boxplot for mbqn-data
plot.new()
mbqnBoxplot(mbqn_mtx,
            add.leg = FALSE, main = "MBQN data",
            cex.lab = 1.2,
            cex = 1.2,
            ylim = ylim)
```

<br /><br />
Example 3: Apply a two-sided t-test before and after application
of different normalizations to a simulated, differentially expressed and
distorted RI feature. The feature is obtained from a simulated dataset
where each sample is distorted in mean and scale.

```{r example3, eval = TRUE}
#plot.new()
mtx <- mbqnSimuData("omics.dep", show.fig = FALSE)
mtx.mod <- mbqnSimuDistortion(mtx, s.mean = 0.05, s.scale = 0.01)
mtx2 <- mtx.mod
mtx.mod <- mtx.mod$x.mod

res <- mbqnGetNRIfeatures(mtx.mod, low_thr = 0.5)

# undistorted feature
feature1 <- mtx[1,]
# distorted feature
feature1mod = mtx.mod[1,]
# feature after normalization
qn.feature1 = mbqn(mtx.mod, verbose = FALSE)[1,]
qn.mtx = mbqn(mtx.mod,verbose = FALSE)

mbqn.mtx = mbqn(mtx.mod, FUN = "mean",verbose = FALSE)
mbqn.feature1 = mbqn(mtx.mod, FUN = "mean",verbose = FALSE)[1,]

# Apply t-test:
# undistorted feature
ttest.res0 <- t.test(feature1[seq_len(9)], feature1[c(10:18)],
                    var.equal =TRUE)
# distorted feature
ttest.res1 <- t.test(feature1mod[seq_len(9)], feature1mod[c(10:18)],
                    var.equal =TRUE)
# mbqn normalized distorted feature
ttest.res <- t.test(mbqn.feature1[seq_len(9)], mbqn.feature1[c(10:18)],
                    var.equal =TRUE)

# compare qn, mbqn and original feature
```
```{r figure8, fig.height = 5, fig.width = 6, fig.align = "left", fig.cap = "Fig. 8 "}
plot.new()
mbqnBoxplot(mtx,ylim = c(25,36),
            vals = data.frame(RI = mtx[1,],
                            NRI = mtx[as.numeric(names(res$nri)[2]),]),
            y.intersp= 1.5,
            main = "Undistorted data")
```
```{r figure9, fig.height = 5, fig.width = 6, fig.align = "left", fig.cap = "Fig. 9 "}
plot.new()
mbqnBoxplot(mtx.mod,ylim = c(25,39),
            vals = data.frame(RI = mtx.mod[1,],
                    NRI = mtx.mod[as.numeric(names(res$nri)[2]),]),
            y.intersp= 1.5,
            main = "Distorted data")
```
```{r figure10, fig.height = 5, fig.width = 6, fig.align = "left", fig.cap = "Fig. 10 "}
plot.new()
mbqnBoxplot(qn.mtx,ylim = c(25,39),
            vals = data.frame(RI = qn.mtx[1,],
                    NRI = qn.mtx[as.numeric(names(res$nri)[2]),]),
            y.intersp= 1.5,
            main = "QN data")
```
```{r figure11, fig.height = 5, fig.width = 6, fig.align = "left", fig.cap = "Fig. 11 "}
plot.new()
mbqnBoxplot(mbqn.mtx,ylim = c(25,39),
            vals = data.frame(RI = mbqn.mtx[1,],
                    NRI = mbqn.mtx[as.numeric(names(res$nri)[2]),]),
            y.intersp= 1.5,
            main = "MBQN data")
```
```{r figure12, fig.height = 5, fig.width = 6, fig.align = "left", fig.cap = "Fig. 12 "}
plot.new()
matplot(t(rbind(feature1 = feature1,
    feature1.mod = (feature1mod-mean(feature1mod))/25+mean(feature1),
    qn.feature1 = (qn.feature1-mean(qn.feature1))+mean(feature1),
    mbqn.feature1 = (
        mbqn.feature1-mean(mbqn.feature1))+mean(feature1))),
    type = "b", lty = c(1,1,1), pch = "o",
    ylab = "intensity",
    xlab = "sample",
    main = "Differentially expressed RI feature",
    ylim = c(34.48,34.85))
legend(x=11,y= 34.86, legend = c("feature","distorted feature/25" ,
                                "QN feature", " MBQN feature"),pch = 1,
        col = c(1,2,3,4), lty= c(1,1,1,1), bty = "n", y.intersp = 1.5,
        x.intersp = 0.2)
legend(x = .1, y = 34.6,
        legend = paste("p-value (t-test) =",round(ttest.res1$p.value,2),
            "\np-value (t-test, mbqn) =", round(ttest.res$p.value,4)),
        bty = "n", x.intersp = 0)


if (ttest.res$p.value<0.05)
    message("H0 (=equal mean) is rejected!")


# print(mtx2$x.mod)
# print(mtx2$mx.offset)
# print(mtx2$mx.scale)
# print(paste("ttest.undistorted =",ttest.res0)) 
# print(paste("ttest.distorted =", ttest.res1))
# print(paste("ttest.mbqndistorted =", ttest.res))
```

<br /><br />
Example 4: This example downloads an LFQ intensity dataset from the PRIDE 
repository, normalizes the data, identifies RI/NRI features, and give graphical 
output. One can choose between four data sets. By default data files are stored 
in the current working directory currentdir/PXDxxx.

The following example illustrates normalization of LFQ intensity data from the PRIDE repository.
For illustration purpose, we use the data set "PXD001584" (file size 10.8 MB)

Other data examples are <br />
* "PXD005138" - contains one RI feature (file size 7440.6 MB) <br />
* "PXD005861" - contains one RI feature (file size 334.7 MB) <br />
* "PXD006617" - contains one RI feature (file size 290.8 MB) <br />

```{r example4, eval = TRUE}
library("SummarizedExperiment")
pxd_id <- "PXD001584"   

# Download file from PRIDE to currentdir/PXDxxx
# getPXDfile(pxd_id)

# Load file
out <- mbqnLoadFile(pxd_id, file.pattern = "proteinGroups.txt")

# filter for potential contaminants and identified only by site features
out <- out[!rowData(out)[["ixs"]],] 

# extract data and feature annotation
mtx <- assays(out)[["data"]]
featureAnnotations <- rowData(out)

low_thr <- 0.5
ylim <- NULL
ix <- seq_len(ncol(mtx))

ix <- c(1:9,19:27)
mtx <- mtx[,ix]
ylim.qn <- ylim <- c(22.5,36)

####################################################################################
```
```{r figure13, fig.height = 3, fig.width = 6, fig.align = "left", fig.cap = "Fig. 13 "}
plot.new()
res <- mbqnPlotAll(mtx,
                   FUN = median,
                   low_thr = low_thr,
                   las = 2,
                   type = "l",
                   feature_index = NULL,
                   show_nri_only = TRUE,
                   axis.cex = 0.5,
                   y.intersp= 0.5)
#dev.off()

# get protein name of strongest nri/ri feature
nri_max <- as.numeric(names(which.max(res$nri)))
#featureAnnotations$proteinDescription[nri_max]
#featureAnnotations$proteinName[nri_max]
#featureAnnotations$nbPeptides[nri_max]

df <- lapply(seq_len(ncol(featureAnnotations)),function(j) featureAnnotations[[j]][[nri_max]])
names(df)<- names(featureAnnotations)

# truncate long name strings
df$proteinName <-  paste(strtrim(df$proteinName,80),"...")
df$proteinDescription <- paste(strtrim(df$proteinDescription,80),"...")

colnames(mtx) <- gsub("LFQ intensity","",colnames(mtx))
mbqn.mtx <- mbqn(mtx,FUN = median)
qn.mtx <- mbqn(mtx,FUN = NULL)

# Boxplot of QN intensity features, highlight RI/NRI Features
if(length(ylim)==0) {
  ylim <- c(floor(min(range(mbqn.mtx, na.rm = TRUE))),ceiling(max(range(mbqn.mtx, na.rm = TRUE))))
  ylim.qn <- c(floor(min(range(qn.mtx, na.rm = TRUE))),ceiling(max(range(mbqn.mtx, na.rm = TRUE))))
}
```
```{r figure14, fig.height = 5, fig.width = 6, fig.align = "left", fig.cap = "Fig. 14 "}
plot.new()
mbqnBoxplot(mtx = qn.mtx,main = "QN",
            ylab = "LFQ intensity",
            ylim = ylim, xlab = "",las =2,
            irow = c(as.numeric(names(res$nri))), y.intersp = 0.5)

# select a qn feature:
is.full.feature <- which((apply(is.na(mtx),1,sum))==0)[1]
df <- data.frame(QN.feature = qn.mtx[is.full.feature,])
names(df) <- paste("QN feature", is.full.feature)

```
```{r figure15, fig.height = 5, fig.width = 6, fig.align = "left", fig.cap = "Fig. 15 "}
plot.new()
mbqnBoxplot(mtx = mbqn.mtx,main = "MBQN with QN-feature",
            irow = c(as.numeric(names(res$nri)),is.full.feature),
            xlab= "", las =2,
            ylab = "LFQ intensity",
            vals = df,
            ylim = ylim, y.intersp = 0.5)

#dev.off()
# QN of data and balance only NRI/RI features

df <- data.frame(qn.mtx[as.numeric(names(res$nri)),])
if(ncol(df)==1) df <- t(df)
rownames(df) <- paste("QN feature",names(res$nri))
df2 <- data.frame(mtx[as.numeric(names(res$nri)),])
if(ncol(df2)==1) df2 <- t(df2)
rownames(df2) <- paste("unnormal. feature",names(res$nri))
colnames(df) <- colnames(df2)
df <- rbind(df,df2)
df <- as.data.frame(t(df))
```
```{r figure16, fig.height = 5, fig.width = 6, fig.align = "left", fig.cap = "Fig. 16 "}
  mtx.nri <- mbqnNRI(mtx,FUN = median,low_thr = 0.5, verbose = FALSE)

plot.new()
mbqnBoxplot(mtx = mtx.nri,
            irow = as.numeric(names(res$nri)),
            ylim = ylim.qn, xlab= "", las=2,
            ylab = "LFQ intensity",
            vals = df,lwd = 1.,
            main = "QN with RI/NRI balanced",
            cex.axis = 1, cex.lab = .9, cex = .9, y.intersp = 0.5)
```

## Documentation
To view the documentation for MBQN, type in R
```
??MBQN
```

## References
[1] A. Schad and C. Kreutz, MBQN: an R/Bioconductor package for 
mean/median-balanced quantile normalization. In prep. 2019 <br/>
[2] Ritchie, M.E., Phipson, B., Wu, D., Hu, Y., Law, C.W., Shi, W., and Smyth, 
G.K. (2015). limma powers differential expression analyses for RNA-sequencing 
and microarray studies. Nucleic Acids Research 43(7), e47. <br/>
[3] Ben Bolstad (2018). preprocessCore: A collection of pre-processing 
functions. R package version 1.44.0. 
https://github.com/bmbolstad/preprocessCore. <br/>
[4] Laurent Gatto (2019). rpx: R Interface to the ProteomeXchange Repository. 
R package version 1.18.1. https://github.com/lgatto/rpx. 
