---
title: "MBQN vignette"
#date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{MBQN Package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This package contains a modified quantile normalization (QN) for preprocessing and analysis of omics or other matrix-like organized data with intensity values biased by global, columnwise distortions of intensity mean and scale. The modification balances the mean intensity of features (rows) which are rank invariant (RI) or nearly rank invariant (NRI) across samples (columns) before quantile normalization [1]. This helps to prevent an over-correction of the intensity profiles of RI and NRI features by classical QN and therefore supports the reduction of systematics in downstream analyses. Additional package functions help to detect, identify, and visualize potential RI or NRI features in the data and demonstrate the use of the modification.

## Installation
To install this package, you need R version >= 3.6.

For installation from Bioconductor run in R:

```{r bc-installation, eval = FALSE}
# if (!requireNamespace("BiocManager", quietly = TRUE))
#    install.packages("BiocManager")
BiocManager::install("MBQN")
```

## Dependencies
The core of the MBQN package uses normalizeQuantiles() from the package `limma` [2], available at https://bioconductor.org/packages/release/bioc/html/limma.html, for computation of the quantile normalization. Optionally, `normalize.quantiles()` from the package preprocessCore [3], available at https://bioconductor.org/packages/release/bioc/html/preprocessCore.html, can be used. 

The function `getPXDfile()` in MBQN uses data from the PRIDE repository. To run this function one needs the R package `rpx` [4] to download the data. 

To install these packages in R run: 

```{r dependencies1, eval = FALSE}
# if (!requireNamespace("BiocManager", quietly = TRUE))
#    install.packages("BiocManager")
# BiocManager::install(pkgs = c("preprocessCore","limma","rpx","SummarizedExperiment"))
```


## Usage
The package provides two basic functions: `mbqn()` applies QN or mean/median-balanced quantile
normalization (MBQN) to a matrix. 
`mbqnNRI()` applies quantile normalization and mean/median-balanced quantile 
normalization only to selected NRI and RI features, specified by a threshold or manually. 
The input matrix may contain NAs. To run one of these functions you will need to provide an
input matrix (see Examples). The argument `FUN` is used to select between classical quantile
normalization (default), and mean or median balanced quantile normalization. The function
`mbqnGetNRIfeatures()` and `mbqnPlotRI()` can be used to check a data matrix for RI or NRI 
features. They provide a list of potential RI/NRI features together 
with their rank invariance frequency and a graphical output. 

## Examples
Example 1: Generate a distorted omics-like matrix of log2-transformed 
intensities with missing values and a single rank invariant feature:
```{r example1, eval = TRUE}
## basic example
library("MBQN")
set.seed(1234)
# data generation 
mtx <- mbqnSimuData("omics.dep")
# data distortion
mtx <- mbqnSimuDistortion(mtx)$x.mod
```
```{r figure1, fig.height = 5, fig.width = 6, fig.align = "left", fig.cap = "Fig. 1 Boxplot of the unnormalized, distorted intensity data matrix. The first feature is an RI feature (red line). It has maximum intensity for each sample!"}
plot.new()
mbqnBoxplot(mtx, irow = 1, main = "Unnormalized")
```
<br />
Apply check for rank invariant (RI) or nearly rank invariant (NRI) features to 
the data matrix and visualize result:
```{r, eval = TRUE}
res <- mbqnGetNRIfeatures(mtx, low_thr = 0.5)
```
```{r figure2, fig.height = 3, fig.width = 6, fig.align = "left", fig.cap = "Fig. 2 Correctly detected and identified RI feature with a data coverage of 100% across samples."}
plot.new()
mbqnPlotRI(res)
```
<br />
Apply quantile normalization with and without balancing the RI feature and 
compare the intensity features: 
```{r figure3, fig.height = 5, fig.width = 6, fig.align = 'left', fig.cap = "Fig. 3 Quantile normalized intensities with balanced and unbalanced normalized RI feature. Classical quantile normalization suppresses any intensity variation of the RI feature, while the MBQN preserves its variation while reducing systematic batch effects!"}
plot.new()
mbqn.mtx <- mbqnNRI(x = mtx, FUN = median, verbose = FALSE) # MBQN
qn.mtx <- mbqnNRI(x = mtx, FUN = NULL, verbose = FALSE) # QN
mbqnBoxplot(mbqn.mtx, irow = res$ip, vals = data.frame(QN = qn.mtx[res$ip,]), main = "Normalized")
```

<br /><br />
Example 2: Visualize  the effect of normalization on
rank mixing and rank invariant intensity features by comparing the intensity distribution of unnormalized, quantile and mean/median-balanced quantile normalized data on a matrix where rows represent features, e.g. of protein
abundances/intensities, and columns represent samples. 

```{r example2, eval = TRUE}
## basic example
library("MBQN")
set.seed(1234)
mtx <- mbqnSimuData("omics.dep") 
# Alternatively: mtx <- matrix(
#            c(5,2,3,NA,2,4,1,4,2,3,1,4,6,NA,1,3,NA,1,4,3,NA,1,2,3),ncol=4)
```
<br />
Perform QN, median balanced QN, and QN with median balanced NRI feature.
```{r, eval = TRUE}
qn.mtx <- mbqn(mtx,FUN=NULL, verbose = FALSE)
mbqn.mtx <- mbqn(mtx,FUN = "median", verbose = FALSE)
qn.nri.mtx <- mbqnNRI(mtx,FUN = "median", low_thr = 0.5, verbose = FALSE)
```
<br />
Check saturation i.e. for rank invariance.
```{r, eval = TRUE}
res <- mbqnGetNRIfeatures(mtx, low_thr = 0.5)

ylim <- range(mtx, na.rm =TRUE)
```
```{r figure4, fig.height = 5, fig.width = 6, fig.align = "left", fig.cap = "Fig. 4 "}
# create a boxplot for qn-data with nri features median balanced
plot.new()
mbqnBoxplot(qn.nri.mtx, irow = res$ip,
            add.leg = TRUE, main ="QN data with MBQN NRI feature",
            cex.lab = 1.2,
            cex = 1.2,
            ylim = ylim)
```
```{r figure5, fig.height = 5, fig.width = 6, fig.align = "left", fig.cap = "Fig. 5 "}
# create a boxplot for x
plot.new()
mbqnBoxplot(mtx,
            add.leg = FALSE, main = "Data",
            cex.lab = 1.2,
            cex = 1.2,
            ylim = ylim)
```
```{r figure6, fig.height = 5, fig.width = 6, fig.align = "left", fig.cap = "Fig. 6 "}
# create a boxplot for qn-data
plot.new()
mbqnBoxplot(qn.mtx, irow = res$ip,
            add.leg = FALSE, main = "QN data",
            cex.lab = 1.2,
            cex = 1.2,
            ylim = ylim)
```
```{r figure7, fig.height = 5, fig.width = 6, fig.align = "left", fig.cap = "Fig. 7 "}
# create a boxplot for mbqn-data
plot.new()
mbqnBoxplot(mbqn.mtx,
            add.leg = FALSE, main = "MBQN data",
            cex.lab = 1.2,
            cex = 1.2,
            ylim = ylim)
```

## Documentation
To view the documentation for MBQN, type in R
```
??MBQN
```

## References
[1] A. Schad and C. Kreutz, MBQN: an R/Bioconductor package for 
mean/median-balanced quantile normalization. In prep. 2019 <br/>
[2] Ritchie, M.E., Phipson, B., Wu, D., Hu, Y., Law, C.W., Shi, W., and Smyth, 
G.K. (2015). limma powers differential expression analyses for RNA-sequencing 
and microarray studies. Nucleic Acids Research 43(7), e47. <br/>
[3] Ben Bolstad (2018). preprocessCore: A collection of pre-processing 
functions. R package version 1.44.0. 
https://github.com/bmbolstad/preprocessCore. <br/>
[4] Laurent Gatto (2019). rpx: R Interface to the ProteomeXchange Repository. 
R package version 1.18.1. https://github.com/lgatto/rpx. 
